---
title: "Lecture 04. Interdependence"
author: "Alex Shpenev"
date: "1/24/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clustering of observations.

Up until now we were talking about observations as being independent from each other. This is not always the case.

THe ashkenazi dataset on breast cancer:
```{r}
library(asaur)
data(ashkenazi)
ashkenazi[ashkenazi$famID %in% c(1, 9, 94), ]
```

```{r}
library(timereg)
data(diabetes)
head(diabetes)
```



# The notion of frailty

$$ h_{ij}(t_{ij}) = h_0(t){ij}\omega_i e^{z_{ij}\beta} $$

Frailty is usually modeled using a gamma distribution but sometimmes other distributions are used.

$$ g(\omega, \theta) = \frac{\omega^{1/\theta  - 1}e^{-\omega/\theta}}{\Gamma(1/\theta)\theta^{1/\theta}}  $$

An alternative is to assume the following:

$$ \omega_i = e^{\sigma u_i} $$


# A practical example using the Ashkenazi data

Analysis without clustering:
```{r}
result.coxph <- coxph(Surv(age, brcancer) ~ mutant, data=ashkenazi)
summary(result.coxph)
```

Analysis with clustering:

```{r}
result.coxph.cluster <- coxph(Surv(age, brcancer) ~ mutant + cluster(famID), data = ashkenazi)
summary(result.coxph.cluster)
```

Analysis uzing frailty

```{r}
result.coxph.frail <- coxph(Surv(age, brcancer) ~ mutant + frailty(famID), data = ashkenazi)
summary(result.coxph.frail)
```

Another approach using a new package


```{r}
library(coxme)
result.coxme <- coxme(Surv(age, brcancer) ~ mutant + (1|famID), data=ashkenazi)
summary(result.coxme)
```


# Let's look at retinopathy again

```{r}
result.coxme <- coxme(Surv(time, status) ~ treat + as.factor(adult) + treat*as.factor(adult) + (1 | id), data=diabetes)
summary(result.coxme)
```



# Cause specific hazards

Let's look at prostate cancer data
```{r}
data(prostateSurvival)
prostateSurvival <- within(prostateSurvival, {
   status.prost <- as.numeric({status == 1}) 
   status.other <- as.numeric({status == 2})}) 
attach(prostateSurvival)
prostateSurvival.highrisk <- prostateSurvival[{{grade == "poor"} & 
   {stage=="T2"} & {ageGroup == "80+"}},]
head(prostateSurvival.highrisk) 

result.prostate.km <- survfit(Surv(survTime, event=status.prost) ~ 1,
    data=prostateSurvival.highrisk)
result.other.km <- survfit(Surv(survTime, event=status.other) ~ 1,
    data=prostateSurvival.highrisk) 

surv.other.km <- result.other.km$surv 
time.km <- result.other.km$time/12 
surv.prost.km <- result.prostate.km$surv
cumDist.prost.km <- 1 - surv.prost.km
plot(cumDist.prost.km ~ time.km, type="s", ylim=c(0,1), lwd=2, 
    xlab="Years from prostate cancer diagnosis",  col="blue") 
lines(surv.other.km ~ time.km, type="s", col="green", lwd=2) 
layout(matrix(c(2,1,3), ncol=3),
  heights=c(5,5,5), widths=c(0.5, 5, 0.5))
layout.show(3)
par(mar=c(5,2,2,2), cex.axis=1.65, cex.lab=1.65)

#par(mar=c(5, 4, 4, 4) + 0.1)
plot(cumDist.prost.km ~ time.km, type="s", ylim=c(0,1), lwd=2, xlab="Years from prostate cancer diagnosis",
   axes=F, col="blue")
lines(surv.other.km ~ time.km, type="s", col="green", lwd=2)
axis(1)
axis(2, at=seq(0,1,0.2), las=1)
#axis(3)
axis(4, at=seq(0,1,0.2), labels=seq(1,0,-0.2), las=1)
box()
text(20/12, 0.25, label="Death from\nprostate cancer", cex=1.65)
text(20/12, 0.73, label="Death from\nother causes", cex=1.65)

x <- c(0,5)
y <- c(0,0.25)
par(mar=c(0,0,0,0))
plot(x ~ y,type="n",axes=F)
text(x=0.1,y=2.5,"Probability of death from prostate cancer",srt=90,cex=1.65)
plot(x ~ y,type="n",axes=F)
text(x=0.15,y=2.5,"Probability of death from other causes",srt=-90,cex=1.65)
```


# How to account for problems with this graph using the competing risk model


```{r}
rm(status)
rm(survTime)
prostateSurvival.T2 <- prostateSurvival[prostateSurvival$stage =="T2",]
attach(prostateSurvival.T2)
cov.matrix <- model.matrix(~ grade + ageGroup)
head(cov.matrix)
library(cmprsk)
result.prostate.crr <- crr(survTime, status, cov1 = cov.matrix[,-1], failcode = 1)
summary(result.prostate.crr)
```


```{r}
result.other.crr <- crr(survTime, status, cov1 = cov.matrix[,-1], failcode = 2)
summary(result.other.crr)
```

