---
title: "Lecture 5. Multistate models"
author: "Alex Shpenev"
date: "1/25/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reference:
https://cran.r-project.org/web/packages/mstate/vignettes/Tutorial.pdf



# Setting up the data

```{r}
library(mstate)
data(ebmt3)
head(ebmt3)
table(ebmt3$dissub)
```


Setting up the transition matrix

```{r}
tmat <- matrix(NA, 3, 3)
tmat[1, 2:3] <- 1:2
tmat[2, 3] <- 3
dimnames(tmat) <- list(from = c("Tx", "PR", "RelDeath"), to = c("Tx",     "PR", "RelDeath"))
tmat
```

Another way to set up the probability matrix faster:


```{r}
tmat <- transMat(x = list(c(2, 3), c(3), c()), names = c("Tx", "PR", "RelDeath"))
tmat
```

A third way to do the same thing:

```{r}
tmat <- trans.illdeath(names = c("Tx", "PR", "RelDeath"))
tmat
```



What are the possible paths?

```{r}
paths(tmat)
```


# Setting up the data for analysis

```{r}
covs <- c("dissub", "age", "drmatch", "tcd", "prtime")
msbmt <- msprep(time = c(NA, "prtime", "rfstime"), status = c(NA, "prstat", "rfsstat"), data = ebmt3, trans = tmat, keep = covs)
head(msbmt)
```

The number of events in the data:

```{r}
events(msbmt)
```



# Preparing data for regression analysis

```{r}
expcovs <- expand.covs(msbmt, covs[2:3], append = FALSE)
head(expcovs)
```


```{r}
msbmt <- expand.covs(msbmt, covs, append = TRUE, longnames = FALSE)
head(msbmt)
```



# Estimating



```{r}
c1 <- coxph(Surv(Tstart, Tstop, status) ~ dissub1.1 + dissub2.1 +    age1.1 + age2.1 + drmatch.1 + tcd.1 + dissub1.2 + dissub2.2 +    age1.2 + age2.2 + drmatch.2 + tcd.2 + dissub1.3 + dissub2.3 + age1.3 + age2.3 + drmatch.3 + tcd.3 + strata(trans), data = msbmt, method = "breslow")
c1
```



```{r}
msbmt$pr <- 0
msbmt$pr[msbmt$trans == 3] <- 1
c2 <- coxph(Surv(Tstart, Tstop, status) ~ dissub1.1 + dissub2.1 +   age1.1 + age2.1 + drmatch.1 + tcd.1 + dissub1.2 + dissub2.2 +    age1.2 + age2.2 + drmatch.2 + tcd.2 + dissub1.3 + dissub2.3 +    age1.3 + age2.3 + drmatch.3 + tcd.3 + pr + strata(to), data = msbmt,    method = "breslow")
c2
```



# Checking for proportionality

```{r}
cox.zph(c2)
```




# Prediction

```{r}
newd <- data.frame(dissub = rep(0, 3), age = rep(0, 3), drmatch = rep(0, 3), tcd = rep(0, 3), trans = 1:3)
newd$dissub <- factor(newd$dissub, levels = 0:2, labels = levels(ebmt3$dissub))
newd$age <- factor(newd$age, levels = 0:2, labels = levels(ebmt3$age))
newd$drmatch <- factor(newd$drmatch, levels = 0:1, labels = levels(ebmt3$drmatch))
newd$tcd <- factor(newd$tcd, levels = 0:1, labels = levels(ebmt3$tcd))
attr(newd, "trans") <- tmat
class(newd) <- c("msdata", "data.frame")
newd <- expand.covs(newd, covs[1:4], longnames = FALSE)
newd$strata = 1:3
newd
msf1 <- msfit(c1, newdata = newd, trans = tmat)
summary(msf1)
```


# Graph

```{r}
plot(msf1, cols = rep(1, 3), lwd = 2, lty = 1:3, xlab = "Years since transplant",  ylab = "Stratified baseline hazards", legend.pos = c(2, 0.9))
```



# The Aalen-Johansen estimator

```{r}
pt <- probtrans(msf1, predt = 0)
head(pt[[3]])
summary(pt, from = 2)

```



# Graph of the probabilities

```{r}
tmat2 <- transMat(x = list(c(2, 4), c(3), c(), c()))
tmat2
msf1$trans <- tmat2
pt <- probtrans(msf1, predt = 0)
summary(pt, from = 1)
plot(pt, ord = c(2, 3, 4, 1), lwd = 2, xlab = "Days since transplant", ylab = "Prediction probabilities", cex = 0.75, legend = c("Alive in remission, no PR", "Alive in remission, PR", "Relapse or death after PR", "Relapse or death without PR"))
```

